---
title: "ERGM LAB: Exercises"
author: "Zack W Almquist (University of Minnesota)"
date: "May 15, 2017"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE,cache=FALSE}
require(knitr)
knitr::opts_chunk$set(cache.comments=FALSE,
               message=FALSE, warning=FALSE,tidy=TRUE,
                fig.align='center', fig.show='hold', comment = NA, cache=TRUE)
## Helpful functions
logit <- function(p){log(p/(1-p))}
ilogit <- function(x){exp(x)/(1+exp(x))}
```

```{r,results='hide',cache=FALSE}
library(ergm)
library(sna)
library(coda)
```

# Rural Arizona risk Networks

## The Data
The data comes was aggregated by Martina Morris (University of Washington) and Richard Rothenberg (Emory University) and put online at ICPSR. The original data can be found [here](http://www.icpsr.umich.edu/icpsrweb/NAHDAP/studies/22140). In this exercise we are goint to investigate four networks derived from the *Rural Arizona risk networks* in Flagstaff, AZ. These networks were collected from May 1996 to Jan 1998 and originally had 95 respondents interviewed 5 times each. All participants are over 18 years old. This network was generated from random walk sampling strategy. **Warning** These networks are obviously from network sample and not *true* complete networks, but we are going to ignore that for this exercise and will treat them as complete. 

* Name generator
    * Sex, needle, other (illicit) drug contact, social contact in last 6 months
    * Sampling strategy
        * Six seeds chosen at random within same geographic area (Flagstaff) from persons presumed to be at elevated risk for HIV acquisition (through sex and/or drug behaviors)

## Networks

Here, I have constructed 4 networks with attributes: race, ethnicity, sex, and age.

Load the data:
```{r}
load("data/flagstaff_rural.rda")
```

### Needle Network

```{r}
flag_needle_net
plot(flag_needle_net,vertex.col="race",main="Needle")
```

### Sexual Contact Network

```{r}
flag_sex_net
plot(flag_sex_net,vertex.col="race",main="Sex")
```

### Social Contact Network

```{r}
flag_social_net
plot(flag_social_net,vertex.col="race",main="Social")
```

### Drug Contact Network

```{r}
flag_drug_net
plot(flag_drug_net,vertex.col="race",main="Drug")
```

# Exercise

For each network explore the basic properties of the network (e.g., mixing matrix, density, etc.) and then attempt to develop a "good" ERG model for each network. After you have the best ERG model you can construct, use the tutorial to try to interpret the predicted probabilities and odds ratio of a given conditional tie. Further, attempt to explain what these results would mean substantively.


