---
title: "ERGM R Lab"
output:
  html_document:
    toc: false
---

## Step 1

* Install [R](https://cran.r-project.org/), see [R Install](Rinstall.html)
* Install [R-Studio](https://cran.r-project.org/), see [R Install](Rinstall.html)
* Open R-Studio -- See [Set up](Connecting.html)
    * If you want to produce a pdf from th RMARKDOWN you will need to install [LaTex](https://www.latex-project.org/get/)
    * [OSX](http://www.tug.org/mactex/) or [Windows](https://miktex.org/)

## Step 2

* Once you have connected R-Studio to github and downloaded the workshop:
    * Open up the [lab](ERGM_Lab_html.html) 
        * You can also open up the lab Rmd file, [ERGM_lab.Rmd]() in files you downloaded from github
    * Read through the lab and run the R code.
        * Compile the whole lab into a .pdf if you installed latex
    * Do the exercises at the end of the lab (data provided through git).
## Notes
    * This lab requires you to install packages
    ```{r,eval=FALSE}
    install.packages("sna")
    install.packages("network")
    install.packages("ergm")
    install.packages("coda")
    install.packages("knitr")
    ```

* Loading packages in R

    ```{r}
library(ergm)
data(florentine) 
flomarriage 
par(mfrow=c(1,2)) 
plot(flomarriage, main="Florentine Marriage", cex.main=0.8) 
summary(flomarriage~edges) 
```

